Icicle: write once, run once

### 2016-05-24 2016fhpc/section/02-Source.tex
typing rules

The grammar for Icicle is given in figure~\ref{fig:source:grammar}.
Value types are given in $T$ and can be numbers, booleans or maps.
Modal types, $T_M$ can be a pure value type, or a modality associated with a value type.
Function types, $T_\to$, can be either a non-function modal type, or a function from modal type to function type.
As Icicle is a first-order language, function types are not value types.

@@

The typing rules for Icicle are given in figure~\ref{fig:source:type:exp}.
The judgment form $\Typecheck{\Gamma}{\mi{Exp}}{T_M}$ means that under context $\Gamma$, the expression has a given value type and modality.
$\TypecheckP{\mi{Prim}}{T_\to}$ associates a primitive with its function type.
$\TypecheckApp{T_\to}{\ov{T_M}}{T_M}$ is used for automatically boxing and unboxing modalities in function application: a function type applied to a list of argument value types and modalities results in a single value type and modality.
The next two judgments,
$\TypecheckS{\Gamma}{\mi{Fun}}{\Gamma}$ and
$\TypecheckS{\Gamma}{\mi{Exp}}{\Gamma}$
both takes an input environment and function or query, and returns an environment containing the function or query name and its type.
Finally, $\TypecheckS{}{\mi{Top}}{\Gamma}$ takes a top-level definition with a table, functions and queries, and returns a context containing only the types of the queries.

The rule (TcVar) performs variable lookup in the context, while (TcNat) shows that number constants have type $\NN$.
(TcBox) shows that a pure value can be converted to an @Element@ or @Aggregate@, by either replicating it along the stream or returning it at the end of the stream.
The premise $\tau~\in~T$ requires that $\tau$ is a pure value, rather than an @Element@ or an @Aggregate@.

(TcPrimApp) and (TcFunApp) find the result type of primitive and function application.
They first find the type of the primitive or function, and the types of the arguments, then use the application boxing/unboxing judgment $\TypecheckApp{T_\to}{\ov{T_M}}{T_M}$ to find the return type and modality of the application.
The set-comprehension syntax $\{e_i\}_i$ means that this rule allows any number of arguments.

(TcLet) is a standard let rule, finding the type of the bound expression and inserting it into the context when typechecking the remaining expression.

(TcFold) finds the type of a @fold@ expression.
The initial value of the fold must be pure.
The update has the current value of the fold inserted in its context as an @Element@, and the update must be an @Element@.
The final return type of the fold is an @Aggregate@.

(TcFilter) requires its predicate to be a stream of booleans ($@Element@~\BB$).
The rest of the filter must be an @Aggregate@; returning an @Element@ stream with a different rate is not allowed, as explained in\sref{s:IcicleSource:bounded}.
(TcGroup) is similar to filter, except returning a @Map@.

The rules (PrimArith), (PrimRel) and (PrimLookup) give function types for each primitive.

The next two rules show how function application boxes and unboxes modalities.

(AppArgs) applies when a function is applied to exactly the right arguments, and simply returns the result of the function.

(AppRebox) is used when a pure function is applied to a non-pure value.
If the arguments and the result type are all pure, then the function can be applied to arguments of a particular modality, and the return type becomes that modality too.
Note that (TcBox) rule can also apply implicit boxing here; when only some of the arguments are non-pure, the pure ones can be boxed.

(CheckFun) inserts the argument types into the context and returns a new context with the function's type, which is the type of the arguments to the result type.

(CheckQuery) checks a query's types, and makes sure it returns an @Aggregate@.

Finally, (CheckTop) checks a whole top-level program.
It first creates a new context with the table columns bound to @Element@s, and typechecks each function.
Then, each query is checked, with the table columns and all functions available.
The final returned context is all query names and their types.

>>>>

The grammar for Icicle is given in figure~\ref{fig:source:grammar}.
Value types are given in $T$ and can be numbers, booleans or maps.
Modal types, $\TauMode$ can be a pure value type, or a modality associated with a value type.
Function types, $\TauFun$, can be either a non-function modal type, or a function from modal type to function type.
As Icicle is a first-order language, function types are not value types.

@@

The typing rules for Icicle are given in figure~\ref{fig:source:type:exp}.
Judgment $\Typecheck{\Gamma}{\mi{Exp}}{\TauMode}$ means that under context $\Gamma$, the expression has a given value type and modality.
Judgment $\TypecheckP{\mi{Prim}}{\TauFun}$ associates a primitive with its function type.
Judgment $\TypecheckApp{\TauFun}{\ov{\TauMode}}{\TauMode}$ is used for automatically boxing and unboxing modalities in function application: a function type applied to a list of argument value types and modalities results in a single value type and modality.
Judgment $\TypecheckS{\Gamma}{\mi{Def}}{\Gamma}$ takes an input environment and function or query, and returns an environment containing the function or query name and its type.
Finally, $\TypecheckS{}{\mi{Top}}{\Gamma}$ takes a top-level definition with a table, functions and queries, and returns a context containing only the types of the queries.

Rule TcVar performs variable lookup in the context, while TcNat shows that number constants have type $\NN$.

Rule TcBox shows that a pure value can be converted to an @Element@ or @Aggregate@, by either replicating it along the stream or returning it at the end of the stream.
The premise $\tau~\in~T$ requires that $\tau$ is a pure value, rather than an @Element@ or an @Aggregate@.

Rules TcPrimApp and TcFunApp find the result type of primitive and function application.
They first find the type of the primitive or function, and the types of the arguments, then use the application boxing/unboxing judgment $\TypecheckApp{\TauFun}{\ov{\TauMode}}{\TauMode}$ to find the return type and modality of the application.
The set-comprehension syntax $\{e_i\}$ means that this rule allows any number of arguments.

Rule TcLet is a standard let rule, finding the type of the bound expression and inserting it into the context when typechecking the remaining expression.

Rule TcFold finds the type of a @fold@ expression.
The initial value of the fold must be pure.
The update has the current value of the fold inserted in its context as an @Element@, and the update must be an @Element@.
The final return type of the fold is an @Aggregate@.

Rule TcFilter requires its predicate to be a stream of booleans ($@Element@~\BB$).
The rest of the filter must be an @Aggregate@; returning an @Element@ stream with a different rate is not allowed, as explained in\sref{s:IcicleSource:bounded}.
Rule TcGroup is similar to filter, except returning a @Map@.

The rules PrimArith, PrimRel and PrimLookup give function types for each primitive.

The next judgment shows how function application boxes and unboxes modalities.
This judgment takes a function type and all its arguments, and returns the result type of the function.

Rule AppArgs applies when a function is applied to exactly the right arguments, and simply returns the result of the function.

Rule AppRebox is used when a pure function is applied to a non-pure value.
If the arguments and the result type are all pure, then the function can be applied to arguments of a particular modality, and the return type becomes that modality too.
Note that rule TcBox can also apply implicit boxing here; when only some of the arguments are non-pure, the pure ones can be boxed.
These rules are tricky to get right; an earlier version allowed applying a modal argument to a pure argument if the return type fitted that mode, which can lead to giving the results of folds as initial values for folds.

Rule CheckFun inserts the argument types into the context and returns a new context with the function's type, which is the type of the arguments to the result type.

Rule CheckQuery checks a query's types, and makes sure it returns an @Aggregate@.

Finally, rule CheckTop checks a whole top-level program.
It first creates a new context with the table columns bound to @Element@s, and typechecks each function.
Then, each query is checked, with the table columns and all functions available.
The final returned context is all query names and their types.
